Do we need detailed demographic data to forecast plant population responses to climate change?

### 2015-07-13 manuscript/manuscript.Rmd
small update to ms

Do we need detailed demographic data to forecast the impacts of climate change on plant populations?
========================================================================================

### Andrew T. Tredennick[^corrauth] and Peter B. Adler

*Andrew T. Tredennick, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*

*Peter B. Adler, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*

[^corrauth]: Corresponding author: [atredenn@gmail.com](mailto:atredenn@gmail.com)

@@

Abstract
--------
Forecasting future states of populations has taken on new urgency as the rate of climate change increases. Traditional plant population models have limited utility in this regard because they are based on detailed demographic data from small, localized plots. These models are difficult to scale up to spatial scales relevant to land managers that require such forecasts to make decisions. To overcome the data limitations of traditional population models, some have proposed population models based on population level, rather than individual level, data that is much easier to collect over broad spatial scales. Using such models violates a central assumption of ecology: individuals respond to weather, not populations to climate. Here, we test whether this assumption is important when forecasting climate change impacts on four perennial grass species in a semi-arid Montana grassland. We parameterized two population models, one based on inidividual level data with three vital rates and one on an aggregated version of the same data (percent cover), with climate covariates and compared their accuracy, precision, and sensitivity to climate. The individual level model was more accurate and precise than the aggregate level model when predicting out of sample observations. The aggregate level model made countervailing forecasts to 1% climate changes when compared to the individual level model, indicating the aggregate level model produces incorrect forecasts. When comparing climate effects from both models, the aggregate level model tends to "miss" important climate effects from at least one vital rate for each species. It appears there is no short cut to forecasting climate change impacts on plant populations — detailed demographic data is essential. But, forecasts were very uncertain, so we advocate for a focus on new methods to collect demographic data more efficiently across environmental gradients in space and time.

*Keywords*: forecasting, climate change, grassland, integral projection model, population model

>>>>

\begin{centering}
\textbf{\large{Do we need detailed demographic data to forecast the impacts of climate change on plant populations?}}

\textsc{\small{Andrew T. Tredennick\footnote{E-mail: atredenn@gmail.com} and Peter B. Adler}}

\textit{\small{Department of Wildland Resources and the Ecology Center, 5230 Old Main Hill, Utah State University, Logan, Utah 84322-5230 USA}}

\end{centering}

@@

Abstract
--------
Forecasting future states of populations has taken on new urgency as the rate of climate change increases. Traditional plant population models have limited utility in this regard because they are based on detailed demographic data from small, localized plots. These models are difficult to scale up to spatial scales relevant to land managers that require such forecasts to make decisions. To overcome the data limitations of traditional population models, some have proposed population models based on population level, rather than individual level, data that is much easier to collect over broad spatial scales. Using such models violates a central assumption of ecology: individuals respond to weather, not populations to climate. Here, we test whether this assumption is important when forecasting climate change impacts on four perennial grass species in a semi-arid Montana grassland. We parameterized two population models, one based on inidividual level data with three vital rates and one on an aggregated version of the same data (percent cover), and compared their accuracy, precision, and sensitivity to climate. The individual level model was more accurate and precise than the aggregate level model when predicting out of sample observations. The aggregate level model made countervailing forecasts to 1% climate changes when compared to the individual level model, indicating the aggregate level model produces incorrect forecasts. When comparing climate effects from both models, the aggregate level model tends to "miss" important climate effects from at least one vital rate for each species. It appears there is no short cut to forecasting climate change impacts on plant populations — detailed demographic data is essential. But, forecasts were very uncertain, so we advocate for a focus on new methods to collect demographic data more efficiently across environmental gradients in space and time.

*Key words: forecasting, climate change, grassland, integral projection model, population model*

### 2015-07-13 manuscript/manuscript.tex
small update to ms

\section{Do we need detailed demographic data to forecast the impacts of
climate change on plant
populations?}\label{do-we-need-detailed-demographic-data-to-forecast-the-impacts-of-climate-change-on-plant-populations}

\subsubsection[Andrew T. Tredennick and Peter B. Adler]{Andrew T.
Tredennick\footnote{Corresponding author:
  \href{mailto:atredenn@gmail.com}{\href{mailto:atredenn@gmail.com}{atredenn@gmail.com}}}
and Peter B.
Adler}\label{andrew-t.-tredennickcorrauth-and-peter-b.-adler}

\emph{Andrew T. Tredennick, Department of Wildland Resources and the
Ecology Center, Utah State University, Logan, UT}

\emph{Peter B. Adler, Department of Wildland Resources and the Ecology
Center, Utah State University, Logan, UT}

@@

Forecasting future states of populations has taken on new urgency as the
rate of climate change increases. Traditional plant population models
have limited utility in this regard because they are based on detailed
demographic data from small, localized plots. These models are difficult
to scale up to spatial scales relevant to land managers that require
such forecasts to make decisions. To overcome the data limitations of
traditional population models, some have proposed population models
based on population level, rather than individual level, data that is
much easier to collect over broad spatial scales. Using such models
violates a central assumption of ecology: individuals respond to
weather, not populations to climate. Here, we test whether this
assumption is important when forecasting climate change impacts on four
perennial grass species in a semi-arid Montana grassland. We
parameterized two population models, one based on inidividual level data
with three vital rates and one on an aggregated version of the same data
(percent cover), with climate covariates and compared their accuracy,
precision, and sensitivity to climate. The individual level model was
more accurate and precise than the aggregate level model when predicting
out of sample observations. The aggregate level model made
countervailing forecasts to 1\% climate changes when compared to the
individual level model, indicating the aggregate level model produces
incorrect forecasts. When comparing climate effects from both models,
the aggregate level model tends to ``miss'' important climate effects
from at least one vital rate for each species. It appears there is no
short cut to forecasting climate change impacts on plant populations ---
detailed demographic data is essential. But, forecasts were very
uncertain, so we advocate for a focus on new methods to collect
demographic data more efficiently across environmental gradients in
space and time.

\emph{Keywords}: forecasting, climate change, grassland, integral
projection model, population model

>>>>

\begin{centering}
\textbf{\large{Do we need detailed demographic data to forecast the impacts of climate change on plant populations?}}

\textsc{\small{Andrew T. Tredennick\footnote{E-mail: atredenn@gmail.com} and Peter B. Adler}}

\textit{\small{Department of Wildland Resources and the Ecology Center, 5230 Old Main Hill, Utah State University, Logan, Utah 84322-5230 USA}}

\end{centering}

@@

Forecasting future states of populations has taken on new urgency as the
rate of climate change increases. Traditional plant population models
have limited utility in this regard because they are based on detailed
demographic data from small, localized plots. These models are difficult
to scale up to spatial scales relevant to land managers that require
such forecasts to make decisions. To overcome the data limitations of
traditional population models, some have proposed population models
based on population level, rather than individual level, data that is
much easier to collect over broad spatial scales. Using such models
violates a central assumption of ecology: individuals respond to
weather, not populations to climate. Here, we test whether this
assumption is important when forecasting climate change impacts on four
perennial grass species in a semi-arid Montana grassland. We
parameterized two population models, one based on inidividual level data
with three vital rates and one on an aggregated version of the same data
(percent cover), and compared their accuracy, precision, and sensitivity
to climate. The individual level model was more accurate and precise
than the aggregate level model when predicting out of sample
observations. The aggregate level model made countervailing forecasts to
1\% climate changes when compared to the individual level model,
indicating the aggregate level model produces incorrect forecasts. When
comparing climate effects from both models, the aggregate level model
tends to ``miss'' important climate effects from at least one vital rate
for each species. It appears there is no short cut to forecasting
climate change impacts on plant populations --- detailed demographic
data is essential. But, forecasts were very uncertain, so we advocate
for a focus on new methods to collect demographic data more efficiently
across environmental gradients in space and time.

\emph{Key words: forecasting, climate change, grassland, integral
projection model, population model}
